\documentclass[english, a4paper]{article}


\usepackage{amsmath, amssymb, amsthm} % AMS-TEX because of course.
\usepackage{mathtools}        % Load for, amongst other things, paired delimiters.
\usepackage{etoolbox}         % Load to be able to perform conditional tests in macros.
\usepackage{thmtools}         % Load to enable end-of-example marks.
\usepackage{siunitx}          % We sometimes want nicely formatted large numbers and SI units.
\usepackage{polynom}          % Package to help with polynomial long division.
% Import mathematics settings.
\input{preamble_maths}


\usepackage{pgfplots}         % Used for producing pretty graphs of functions.
\usepackage{tikz-3dplot}      % Make fancy 3d plots.
\usepackage{wrapfig}          % Used for producing figures in within the text.
\usepackage{float}            % Force Here placement of floats.
\usepackage{subcaption}       % Used for subfigures.
% Import figure specifics.
\input{preamble_tikz}


\usepackage{enumitem}         % Load for more customisable enumerate environments. Specifically better labels.
\usepackage[hyperfootnotes=false]{hyperref}         % Make the table of contents easier to navigate with hyperlinks.
                                                    % Footnote hyperlinks supressed because otherwise numberless footnotes produce empty anchor errors.
\usepackage{makeidx}          % We want a pretty index to keep track of all the ideas.
\usepackage{tocbibind}        % Include index and bibliography in table of contents.
\usepackage{abstract}         % Load in order to more easily manipulate the appearance of abstracts.
\usepackage{fancyhdr}         % Use to customise the headings to include useful navigation information.
\usepackage{titlesec}         % Use to add 'Lecture' in front of section headings.
\usepackage[titles]{tocloft}  % Use to add same in front of sections in the table of contents.
\usepackage{adjustbox}        % Help us center very wide tables.
% Set lecture presets.
\input{preamble_lecture.tex}


\title{Lecture Notes in Calculus I}
\author{Lectures by Jakob Streipel\thanks{\href{mailto:jakob.streipel@lnu.se}{jakob.streipel@lnu.se}}}
\date{}

\begin{document}
% Use Roman numerals for page numbers for frontmatter.
\pagenumbering{roman}

\maketitle

\begin{abstract}
	\noindent
	These lecture notes are based on the first few chapters of Robert A. Adams's \emph{Calculus: A Complete Course}, \cite{Adams2013}, wherein recommended exercises are also found. All the material covered can be found in there, though the exposition might sometimes be altered. Occasionally there will be small sections marked as `exercise'. These are examples or simpler proofs which are left for the student to think about on their own.  \blfootnote{Last updated \today.}

	\vspace{1em}

	\noindent
	Throughout this document, $\qed$ signifies end proof, and $\blacktriangle$ signifies end of example.
\end{abstract}

\tableofcontents

% Begin mainmatter, and therefore use Arabic page numbering.
\cleardoublepage
\pagenumbering{arabic}

% Start at lecture 0, since the first lecture is preliminary.
\setcounter{section}{-1}

\lecture[January 16, 2017]{Distance and Functions}

\input{lectures/lecture0}


\lecture[January 17, 2017]{Limits}

\input{lectures/lecture1}


\lecture[January 24, 2017]{Continuity}\label{lec2:continuity}

\input{lectures/lecture2}


\lecture[January 26, 2017]{The Derivative}

\input{lectures/lecture3}

\lecture[January 30, 2017]{The Mean-Value Theorem}

\input{lectures/lecture4}


\lecture[February 2, 2017]{The Natural Logarithm}

\input{lectures/lecture5}


\lecture[February 6, 2017]{The Number $e$ and L'H\^{o}pital's Rules}

\input{lectures/lecture6}


\lecture[February 9, 2017]{Sketching Functions}

\input{lectures/lecture7}


\lecture[February 13, 2017]{Overdue Proofs and Antiderivatives}

\input{lectures/lecture8}


\lecture[February 16, 2017]{The Integral}

\input{lectures/lecture9}


\lecture[February 20, 2017]{The Fundamental Theorem of Calculus}

\input{lectures/lecture10}


\lecture[February 23, 2017]{Method of Substitution}

\input{lectures/lecture11}


\lecture[February 27, 2017]{Integration by Parts}

\input{lectures/lecture12}


\lecture[March 2, 2017]{Computing Volumes}

\input{lectures/lecture13}


\lecture[March 6, 2017]{Differential Equations}\label{lec14:differentialequations}

\input{lectures/lecture14}


\lecture[March 9, 2017]{Differential Equations, continued}

\input{lectures/lecture15}



% List of references.
\clearpage
\begin{thebibliography}{99999}
	\bibitem[Ad13]{Adams2013} R.\,A. Adams.
	\emph{Calculus: A Complete Course}.
	Prentice-Hall,
	1136 pages,
	8th revised edition,
	2013.

	\bibitem[Ni99]{Nicholson1999} W.\,K. Nicholson.
	\emph{Introduction to Abstract Algebra}.
	Wiley-Interscience,
	599 pages,
	2nd edition,
	1999. 
\end{thebibliography}

% Make a pretty index.
\clearpage
\printindex

\end{document}
